---
title: "Untitled"
author: "Shiyuan"
date: "2025-05-24"
output: html_document
---
# I. Research Background
Rising atmospheric CO₂ levels due to anthropogenic activities have significant impacts on terrestrial ecosystems, particularly on soil carbon (C) pools. Soil organic carbon (SOC) is a key component of the global carbon cycle, influencing soil fertility, ecosystem productivity, and carbon sequestration potential. Numerous field experiments worldwide have examined how elevated CO₂ affects SOC, but results are variable due to differences in experimental design, duration, soil type, vegetation, and climatic conditions. To synthesize these findings, meta-analytical approaches such as log response ratios (LRR) are often used, allowing standardized comparisons across experiments and identifying general patterns of CO₂ enrichment effects on SOC over time.

# II. Research Purpose
This study aims to aggregate data from multiple CO₂ enrichment experiments to quantify the effects of elevated CO₂ on soil organic carbon (SOC) content across different experimental durations. Specifically, it seeks to calculate log response ratios (LRR) of SOC under elevated vs. ambient CO₂ and analyze how these effects vary by experimental phase: early (<3 years), mid (3–6 years), and late (>6 years).

# 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading required libraries
library(dplyr)
library(lubridate)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(knitr)
library(here)
library(janitor)
```

```{r cars}
summary(cars)

data <- read_csv(here("data_raw", "1249534s1.csv"), skip = 3)
data_clean <- data %>% clean_names()
data_clean <- data_clean %>%
  mutate(sample_date_clean = parse_date_time(sample_date, orders = c("b Y", "m Y", "Y")))

data_selected <- data_clean %>%
  select(experiment, citation, depth, sample_date, time_years,
         ambient_co2_6, increased_co2_7,
         ambient_co2_8, increased_co2_9)



data_lrr <- data_selected %>%
  filter(!is.na(ambient_co2_6), !is.na(increased_co2_7), !is.na(time_years)) %>%
  mutate(
    rr = increased_co2_7 / ambient_co2_6,
    lrr = log(rr),
    phase = case_when(
      time_years < 3 ~ "Early",
      time_years >= 3 & time_years <= 6 ~ "Mid",
      time_years > 6 ~ "Late",
      TRUE ~ NA_character_
    )
  )
summary_lrr <- data_lrr %>%
  group_by(experiment, phase) %>%
  summarise(
    mean_rr = mean(rr, na.rm = TRUE),
    mean_lrr = mean(lrr, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(summary_lrr, caption = "Mean response ratios and log response ratios grouped by experiment and stage")

data_lrr <- data_selected %>%
  filter(
    !is.na(ambient_co2_6), !is.na(increased_co2_7),
    !is.na(ambient_co2_8), !is.na(increased_co2_9),
    !is.na(time_years)
  ) %>%
  mutate(
    rr_6_7 = increased_co2_7 / ambient_co2_6,
    lrr_6_7 = log(rr_6_7),
    rr_8_9 = increased_co2_9 / ambient_co2_8,
    lrr_8_9 = log(rr_8_9),
    phase = case_when(
      time_years < 3 ~ "Early",
      time_years >= 3 & time_years <= 6 ~ "Mid",
      time_years > 6 ~ "Late",
      TRUE ~ NA_character_
    )
  )
summary_lrr <- data_lrr %>%
  group_by(experiment, phase) %>%
  summarise(
    mean_rr_6_7 = mean(rr_6_7, na.rm = TRUE),
    mean_lrr_6_7 = mean(lrr_6_7, na.rm = TRUE),
    mean_rr_8_9 = mean(rr_8_9, na.rm = TRUE),
    mean_lrr_8_9 = mean(lrr_8_9, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(summary_lrr, caption = "
Mean response ratios and log response ratios grouped by experiment and stage")


```

# IV. CONCLUSIONS AND DISCUSSION 
The combined analyses indicate that elevated CO₂ generally increases SOC, but the magnitude of this effect varies across experimental phases. In the early phase (<3 years), the SOC response was typically small or erratic, reflecting initial adjustments to the system. In the middle phase (3-6 years), positive effects became more pronounced as plant inputs and microbial processes adapted to the new CO₂ regime. However, in the late phase (>6 years), the response may plateau or even decline, suggesting that the system is saturated or there are feedback limitations (e.g., nutrient limitations). These findings emphasise the importance of long-term experiments for understanding the effects of CO2 enrichment on soil carbon pools and caution against extrapolating short-term results to long-term ecosystem projections. Future research should further explore potential mechanisms regulating SOC response to elevated CO₂, including nutrient cycling and microbial adaptation.
